title

### 2025-02-26 index.qmd
long overdue commit.

---
title: title
subtitle: subtitle
shorttitle: A Short title # aog-article-pdf
shortauthors: Isaiah      # aog-article-pdf
authors:
  - name: Isaiah Espinoza
    email: gespinoz@umd.edu
    roles: author
    affiliation:
      - name: University of Maryland
        department: Department of Government and Politics
        city: College Park
        state: MD
        country: US
reference-section-title: References
date: "2024-10-16"
date-modified: today
crossref:
  custom: 
    - key: hyp
      kind: float
      reference-prefix: H
      latex-env: hypothesis # required for PDF format
---

@@

# load packages
library(tidyverse)

# load data
load("data/research-paper-data-20241029.Rdata")

# load data dictionary/codebook
codebook <- read.csv("data/survey_codebook.csv")
data_dict <- read.csv("data/data_dictionary.csv")

# also load the raw data dict for comparison
raw_spss_dict <- read.csv("data/raw_spss_dict.csv")
```

```{r}
#| label: post_processing
#| include: false

# Some post-processing of the dataframe :::::::::::::::::::::::::::::::::::::###

# code to recode levels "-99" in factor vars
data <- data |> 
  mutate(across(where(is.factor), ~forcats::fct_recode(., NULL = "-99"))) |> 
  # Reverse response options to show increase in *less concern*
  mutate(q25 = forcats::fct_rev(q25))

# recode factor levels 
data <- data |> 
  # use `pewmethods` to preserve order of factor levels
  mutate(across(c(q19, q20, q21, q22, q23,q24, q25, q26,
                  q30, q31, q32, q33, q34, q35, q36, q37), 
                ~pewmethods::fct_case_when(
    .x == "Not at all confident" |
      .x == "Not at all committed" ~ "Not at all",
    .x == "Not too confident" |
      .x == "Not too committed"|
      .x == "Not too concerned"|
      .x == "Somewhat unconcerned" ~ "Not that much",
    .x == "Somewhat confident" |
      .x == "Somewhat committed" ~ "Somewhat",
    .x == "Very confident" |
      .x == "Very committed" ~ "Very",
    TRUE ~ NA
  )))

# re-assign var labels to variables
# To quickly assign the variable labels, first create a named vector via
# deframe() with values as the variable labels and names as the variable names.
data_labels <- data_dict |>
  select(var, label_var) |> 
  deframe()

# Now assign the labels using the splice operator. Using the splice operator,
# labels are assigned via matching against the variable name, which means that
# variable order does not matter.
data <- data |> 
  labelled::set_variable_labels(!!!data_labels)

```{r}
#| label: fns
#| include: false

source(here::here('src', 'funs.R'))

@@

Even though election officials can do a lot to secure election integrity, there's not much they can do to cement public confidence after election night passes. At best, election officials can ease public anxieties prior to election night.

@@

Supposing such is the case, we can expect that information about *who* (i.e., which groups) election officials are targeting in publicized recruitment efforts would lessen particular election anxieties. That is to say, it is reasonable to expect that telling people *who* will be working as election staff and volunteers would ease election anxieties, therefore improve confidence that the election will be conducted fairly, accurately, and safe for all involved.

In this paper, I report results from a recent survey experiment administered to test whether publicized efforts to recruit veterans to work as election staff and volunteers would improve public trust in elections and ease election anxieties. Results of the survey experiment support the hypothesis that emphasizing veterans as the target of election worker recruitment efforts eases pre-elections anxieties to an extent. In particular, among those in the treatment condition, confidence that the elections would be fair and accurate was higher in comparison to those in the control. In addition, expectations of electoral fraud were lower, as were concerns for voter safety compared to those who read a control vignette. Notable is that there was a significant difference in confidence, and lower expectations of fraud, among those in the treatment condition who believe that the 2020 election was illegitimate.

[Insert why this is an important contribution]

This paper is structured as follows. First, I provide a brief background on public trust in election administration. Second, I synthesize a review of relevant literature with a focus on how political and social science has conceptualized and ascertained public trust in elections. Next, I supply reasoning for why military veterans are singled out as the relevant subset of the population in this study.

-   when distinguished by partisanship...
-   moreover, broken down by other demographic variables such as race, gender, education, etc.
-   holding these factors constant demonstrates a \[increased/decreased\] likelihood that a respondent's confidence would likely \[increase/decrease\], expectations of fraud would likely \[increase/decrease\], and concerns for safety would lessen among those in the treatment group compared to those in the control. These results allow us to attribute the differences to the treatment stimulus, namely, the prospect of having veterans serve as election staff or volunteers.

@@

Election administration is just one institution a part of the larger set of institutions which form the electoral system. As such, the performance of the institution along with the rest "...lends credibility to the outcome of an election: whether it is considered by citizens and the international community to be fair and legitimate." [@stewart2022, 236]. Intuitively, enhancing public trust in elections would best be accomplished by enhancing the *trustworthiness* of the institution, i.e., consistently doing the things that election officials already regularly do come election time. However, trust and confidence in elections has become ever more precarious over the last few election cycles. Especially considering public polling data since 2000 shows that confidence that votes were, or would be, counted as intended was in a consistent decline despite efforts towards bolstering election integrity and trustworthiness [@sances2015]. This is even more pronounced considering the role that partisanship has had on such confidence over accuracy of vote count [@sances2015; @stewart2022].

@@

As such, it is questionable whether we can characterize public trust in the pre-election period as the same trust after the election results have come out. The former is *anticipatory*---i.e., the kind that is more or less anxious given the uncertainties surrounding the election. The latter is *empirical*---a judgement discerned in hindsight after the experience of the election event has occurred. Evaluations of one's own trust and confidence in election administration based on the experience of voting is influenced by that very experience. As discussed by @stewart2022, not only was confidence in vote count accuracy influenced by the voter experience, but so too were evaluations of election officials [-@stewart2022, 242-243]. Not to mention the influence that the election results would also have on such evaluations. In other words, evaluations of trust and confidence post-election are influenced by one's interpretation of their experience. The subtle difference is simply the degree of uncertainty one feels in anticipation of the next election event. This study focuses primarily on that *anticipatory* kind of trust and confidence, which speaks more to those anxieties[^3] based on perceptions of the institution's trustworthiness than upon the particular voter experience.

@@

## Design and Measures

To test the theory that publicized efforts to recruit veterans to work as election staff and volunteers would improve trust in elections and ease election anxieties, a recent survey experiment was embedded in a survey developed and conducted by the Center for Democracy and Civic Engagement (CDCE) at the University of Maryland. The survey was fielded from August 29th, 2024 to September 18th, 2024 on a non-probability sample of 1,287 U.S. citizens 18 years of age or older. Respondents were randomly split into either the treatment (n = 650, 50.51%) control (n = 637, 49.49%) conditions[^Demographic breakdown of the sample are included in Appendix B.].

Survey participants read either a treatment or control vignette, which was a fabricated news article about efforts in Maricopa County, AZ to recruit election staff and volunteers for the 2024 general election. The treatment vignette referred to a program designed to recruit veterans and their family members and describes an interviewee "Jordan Braxton" as an Army veteran. The control vignette simply omitted any mention of veterans and their family members, and didn't describe "Jordan Braxton" as an Army veteran. Beyond those small differences and the headlines, the article vignettes are identical. Any effects can be attributed to the information about veterans in the treatment vignette[^5].

It should be noted that Maricopa County, AZ was chosen as the setting of the story in the vignettes due to the increased scrutiny levied toward election administration there after the 2020 election [@giles2021; @maricopacountyelectionsdepartment2022]. A block of survey items asked specifically about Maricopa County, AZ, followed by an identical block of items that asked the same questions about one's local area. Results between the items where the setting differs are compared and discussed.

### Trust in Elections Admin

@@

Note that, except for the item assessing voter confidence (i.e., votes counted as voters intend), most other question items from the 2022 SPAE inquire into the voter experience after the election has occurred. The items in this study, however, all inquire into confidence over one's expectations in anticipation of upcoming elections. In addition, other survey questionnaires ask similar questions that specifically contrast between a respondent's own local community, their state, and nationwide. This is in addition, or complementary to, items that distinguish between local area officials, state election officials, state government, and the U.S. federal government. The survey of this study, instead, focused on attitudes regarding election workers such as election officials, staff, and volunteers. As such, the items are utilized to comprise a component[^7] I broadly refer to as *trust in elections*.

[^7]: The methodological concerns between the use of (unit weighted) sum scores compared to the use of estimated factor scores for Likert scales composed of multiple Likert items are legion. The use of either computed sum scores or estimated factor scores as a general practice is continuously debated [@mcneish2020; @widaman2022; @mcneish2023; @widaman2024]. The use of either appears to largely depend on the researcher's particular research objective, theoretical suppositions, and necessity. Another common concern is the inappropriate treatment of ordinal variables as interval which levy strong assumptions, e.g., each item contributes equally to the variable being measured and that ordinal response categories are equidistant as though on an interval scale [@wang2013; for contrast, see @robitzsch2020].

A set of five survey items measured public trust in elections by inquiring into the degree of confidence that votes will be counted as voters intend, confidence that electoral systems are secure from technological threats, perceived commitment of election staff, confidence that outcomes will be fair, as well as confidence that the voting process will be fair. The items measuring trust and confidence in elections include:

-   How confident are you that votes in Maricopa County, AZ will be counted as voters intend in the elections this November?[^8]
-   How confident are you that election officials, their staff, and volunteers in Maricopa County, AZ will do a good job conducting the elections this November?
-   Think about the election staff and volunteers who handle the administration and conduct of elections in Maricopa County, AZ. How committed do you think they will be to making sure the elections held this November are fair and accurate?
-   How confident are you that the voting process will be fair in Maricopa County, AZ?
-   How confident are you that election systems in Maricopa County, AZ will be secure from hacking and other technological threats?

[^8]: Note that confidence, or lack thereof, that votes will be counted as voters intend could easily be construed as an expectation of electoral fraud. It is a given that the votes will be counted, but the question is whether they will be counted *as intended*. Although fraud isn't provided as the reason for why votes wouldn't be counted as intended, it also isn't implied that miscounts would be due to some negligent mistake. A nefarious insinuation might be somewhat implicit, but perhaps only to those who are inclined to suspect as such. A lack of confidence here doesn't necessarily impart the reason for that lack of confidence, which makes it suitable as a survey item to measure the presence of trust in election administration rather than an expectation of fraud.

Each item in this series presented four response options from "Not at all [confident/committed]", "Not too [confident/committed]", "Somewhat [confident/committed]", and "Very [confident/committed]". In total, six survey items measure different aspects of trust and confidence in election administration. From these items, I generated a composite score for each respondent that represents their variable level of trust in elections.

### Expecatation of Electoral Fraud

A series of five survey items captured *expectation of electoral fraud*. These items were prefaced with the question, "How likely do you think any or all of the following will happen during this year´s elections in Maricopa County, AZ?" Each item presented four response options: "Not likely at all", "Not too likely", "Somewhat likely", and "Very likely".

-   There will be voter fraud, that is, people who are not eligible to vote will vote, or vote more than once
-   Many votes will not actually be counted
-   Many people will show up to vote and be told they are not eligible
-   A foreign country will tamper with the votes cast in this area to change the results
-   Election officials in Maricopa County, Arizona will try to discourage some people from voting

Similar to the items capturing trust in elections, I computed a composite score for each respondent reflecting their expectations of electoral fraud.

## Model and Data Specification

The outcome of interest concerns both public trust in election administration and the expectation of electoral fraud. Because these variables are assessed by multiple Likert items, conventional methods call for combining the items of each set into separate Likert scales, also known as a summated rating scales.
However, many of the items comprising the trust scale directly oppose items capturing expectations of electoral fraud. For instance, one item has the respondent report their confidence that votes will be counted as voter's intend; whereas a contrasting item has the the respondent report their expectation that many votes will not be counted. These two items directly oppose each other, but there's no guarantee that individual's responded as such.

Since many of the items are not symmetric---i.e., response options are not bipolar with contrasting positive and negative response options---then a "trust in elections" scale could only be understood as a measure of trust in the positive use of the term (i.e., unipolar). Likewise, the expectation of fraud items fit to a scale that corresponds to a unipolar measure of *confident distrust* so to speak.

I would expect composite sum or mean scores, and item responses, on these two measures to be inversely correlated among the sample, but not mutually exclusive. Indeed, this is what I find.

I first constructed two multi-item composite scales of both *trust in elections* and *expectation of fraud*. Cronbach's internal consistency reliability coefficient \alpha were estimated for both scales and were adequate (\alpha > 0.8); the trust in elections scale \alpha = 0.93 and the expectation of electoral fraud scale \alpha = 0.86. In other words, the items in each of the scales were intercorrelated well enough to compose a single scale of a (presumed) unidimensional construct. I then plotted separate histograms of scores across the sample to visualize the distribution for each seen in @fig-dist. Note that for both scales, higher values indicate less expectation of electoral fraud and greater trust in elections, respectively.

```{r}
#| label: fig-dist
#| fig-cap: "Histograms of Composite (mean) Scores for Trust in Elections And Expectation of Electoral Fraud"

# trust scores are not normally distributed, bi-modal, neg skew
trst.scores.hist <- ggplot(data, aes(trst.mean.scores))+
  geom_histogram(aes(y = after_stat(density)), bins = 15,
                 color = 1, fill = "grey")+
  geom_density(lwd = 1.5, color = 1, fill = 1, alpha = 0.35)+
  geom_segment(x = round(mean(data$trst.mean.scores), 2), y = 0, yend = 0.80, color = "red")+
  labs(
    x = "Trust in Elections Mean Scores",
    y = "Density",
    caption = "mean = 3.09, sd = 0.69, median = 3.0, skew = -0.67, kurtosis = 0.25"
  )+
  annotate("text", x = round(mean(data$trst.mean.scores), 2), y = 0.85,
           label = paste("mean =", round(mean(data$trst.mean.scores), 2)), color = "red")

# eef scores are not normally distributed, also bi-modal pos skew
eef.scores.hist <- ggplot(data, aes(eef.mean.scores))+
  geom_histogram(aes(y = after_stat(density)), bins = 15,
                 color = 1, fill = "grey")+
  geom_density(lwd = 1.5, color = 1, fill = 1, alpha = 0.35)+
  geom_segment(x = round(mean(data$eef.mean.scores), 2), y = 0, yend = 0.80, color = "red")+
  labs(
    x = "Exp. of Electoral Fraud Mean Scores",
    y = "Density",
    caption = "mean = 2.27, sd = 0.75, median = 2.2, skew = 0.19, kurtosis = -0.55"
  )+
  annotate("text", x = round(mean(data$eef.mean.scores), 2), y = 0.85,
           label = paste("mean =", round(mean(data$eef.mean.scores), 2)), color = "red")

# place plots side by side
gridExtra::grid.arrange(eef.scores.hist, trst.scores.hist, ncol = 2)
```

Given the non-normal (or at least un-equal skew of the) distribution of the two scale scores and ordinal nature of the variable items, I conducted a Spearman's rank correlation test [@spearman1907] and found a negative correlation of $\rho$ -0.49 (Kendall's \tau = -0.39) between summated scores on the two scales.

The negative correlation between the items and the composite scores makes sense. Theoretically speaking, a person who is certain that electoral fraud will occur clearly lacks trust in the election. That is, expectation of electoral fraud does imply lack of trust in elections. However, the reverse isn't also necessarily true; it is reasonable to hold a degree of trust in elections alongside expectations that electoral fraud will occur. Holding both considerations isn't necessarily an example of cognitive dissonance, as a lack of confidence in elections doesn't necessarily imply an expectation of electoral fraud. A lack of trust in elections might render one open to potentialities, but expectation of fraud is in no way necessitated by lacking trust in elections. When considered along a spectrum, both can be considered as degrees of *confidence*; and in the absence of confidence, insecure.

Accordingly, those whose responses to the "trust in elections" items would be cancelled out by their responses to the "expectation of fraud" items. Those who are, or are close to being, equally confident in both directions would be the most *insecure* about their expectations[^Insecure in terms of their confidence. Not insecure as to imply some anxious or concerned affect.]. Having no trust in elections and no expectation that electoral fraud will occur is functionally equivalent to having equal degree of both. Quite literally, an uncertain feeling of "it could go one way or the other". Operationally, this would reflect a middling 0 on a scale of *confidence*. Seeing as how expectation of electoral fraud is also a level of confidence, then it is reasonable to place all items onto a symmetric scale where positive values represent trust in elections and negative values represent expectation of electoral fraud.

The primary dependent variable is *confidence in elections*, which reflects one's trust in election administration minus their expectations of electoral fraud. Operationally, the dependent variable is a linear composite scale of item responses across items measuring trust in elections and expectations of electoral fraud re-scaled to range from -3 to 3. Items comprising trust in elections were positively coded to range from 0 "Not at all confident/committed" to 3 "Very confident/committed". Likewise, items capturing expectation of electoral fraud were coded negatively to range from 0 "Not likely at all" to -3 "Very likely".

Computed mean or sum scores for multiple Likert items are commonly acceptable, as they are relatively robust in spite of violating assumptions of a normal distribution of data (or residuals).
due to the somewhat non-normal distribution of the sum or mean scores and the ordinal nature of the variable items, an ordered regression analysis is preferred.

The primary independent variable is the prospect of military veterans as election workers (i.e., staff or volunteers).

@@

The key question is whether trust in elections (or voter confidence) is response to the treatment vignette.

Note that the question wording for the survey items derived from the 2022 SPAE are modified slightly but substantially. The question wording in this study is most similar to those in 2022 SPAE that assess confidence that votes will be counted as voters intend without specific reference to their own vote (e.g., in their county, state, or nationwide). In this study, two survey items ask about confidence in vote counts in one specific county in Arizona and another referencing their local community. Moreover, the 2022 SPAE includes "I don't know" as a response option, whereas the comparable item in this study did not include a "don't know" response option. Comparison between responses to the two similar items was conducted....

>>>>

---
title: title
subtitle: subtitle
shorttitle: A Short title # aog-article-pdf
shortauthors: Isaiah      # aog-article-pdf
authors:
  - name: Isaiah Espinoza
    email: gespinoz@umd.edu
    roles: author
    affiliation:
      - name: University of Maryland
        department: Department of Government and Politics
        city: College Park
        state: MD
        country: US
reference-section-title: References
date: "2024-10-16"
date-modified: today
knitr: 
  opts_chunk: 
    dev: "ragg_png" # same as knitr::opts_chunk$set(dev = "ragg_png")
---

@@

set.seed(1234)

# load packages
library(tidyverse)
library(easystats)

options("huxtable.latex_use_fontspec" = TRUE) # if TRUE, uses the LaTeX “fontspec” package which lets you use the same fonts in TeX and HTML. You will need to use the xetex or xelatex engine for output.

# load data
load("data/research-paper-df-20250203.Rdata") #df

# load data dictionary/codebook
codebook <- read.csv("data/survey_codebook.csv")
df_dict <- read.csv("data/df_dictionary.csv")

@@

Even though election officials can do a lot to secure election integrity, there's not much they can do to cement public confidence after election night passes. At best, election officials can ease public insecurities prior to election night.

@@

Supposing such is the case, we can expect that information about *who* (i.e., which groups) election officials are targeting in publicized recruitment efforts would lessen particular election insecurity, and concurrently, boost confidence. That is to say, it is reasonable to expect that telling people *who* will be working and volunteering as election staff would ease election insecurity, and therefore improve confidence that the election will be conducted fairly, accurately, and safe for all involved.

In this paper, I report results from a recent survey experiment administered to test whether publicized efforts to recruit veterans to work as election staff and volunteers would improve public trust in elections and ease election insecurity. Results of the survey experiment support the notion that emphasizing veterans as the target of election worker recruitment efforts eases pre-election insecurity to a limited extent. Expectations of electoral fraud and concerns for voter safety were lower among those who read a fabricated announcement that veterans are being recruited to work as election staff and volunteers compared to those who read a control vignette where veterans were not mentioned. Notable is that there was a significant difference in confidence among those in the treatment condition who believe that results of the 2020 election were illegitimate.

This paper is structured as follows. First, I provide a brief background on public trust in election administration. I synthesize a review of relevant literature with a focus on how political and social science has conceptualized and ascertained public trust in elections. Next, I supply reasoning for why military veterans are singled out as the relevant subset of the population in this study. I then explicitly provide the simple theory and testable hypotheses of the study before moving on to describe the design of the survey, the measurement instruments therein, and the conceptual and operational definitions of the variables of interest. I explain my reasoning and method for constructing the primary dependent variable, which I broadly refer to as *confidence in elections*. I preview the different methods of analysis alongside presentation of the results. Before concluding, I dedicate time to discuss the limitations of the study and lessons learned. I close by offering my interpretation of the results and suggest potential avenues for future inquiry.

@@

Elections administration is just one institution a part of the larger set of institutions which form the electoral system. As such, the performance of the institution along with the rest "...lends credibility to the outcome of an election: whether it is considered by citizens and the international community to be fair and legitimate." [@stewart2022, 236]. Intuitively, enhancing public trust in elections would best be accomplished by enhancing the *trustworthiness* of the institution, i.e., consistently doing the things that election officials already regularly do come election time. However, trust and confidence in elections has become ever more precarious over the last few election cycles. Especially considering public polling data since 2000 shows that confidence that votes were, or would be, counted as intended was in a consistent decline despite efforts towards bolstering election integrity and trustworthiness [@sances2015]. This is even more pronounced considering the role that partisanship has had on such confidence over accuracy of vote count [@sances2015; @stewart2022].

@@

As such, it is questionable whether we can characterize public trust in the pre-election period as the same trust after the election results have come out. The former is *anticipatory*---i.e., the kind that is more or less anxious given the uncertainties surrounding the election. The latter is *empirical*---a judgement discerned in hindsight after the experience of the election event has occurred. Evaluations of one's own trust and confidence in election administration based on the experience of voting is influenced by that very experience. As discussed by @stewart2022, not only was confidence in vote count accuracy influenced by the voter experience, but so too were evaluations of election officials [-@stewart2022, 242-243]. Not to mention the influence that the election results would also have on such evaluations. In other words, evaluations of trust and confidence post-election are influenced by one's interpretation of their experience. The subtle difference is simply the degree of uncertainty one feels in anticipation of the next election event. This study focuses primarily on that *anticipatory* kind of confidence, which speaks more to those insecurities[^3] based on perceptions of the institution's trustworthiness than upon the particular voter experience.

@@

## Theory and Hypotheses

Considering the generally positive perception afforded to veterans by the public, it is reasonable to suspect that efforts that promote recruitment of veterans and their family members to work as election staff and volunteers would boost public confidence in election administration. Indeed, that is the primary hypothesis of this study. Although citizens may vote in various ways across the country (e.g., by mail, ballot drop box, in-person), the simple announcement that election officials are engaging in efforts to recruit veterans to work as election staff may boost confidence in elections administration regardless of how, or whether, an individual plans to vote. Formally,

> H~1~: Announcements that election officials are recruiting miilitary service veterans to work as election staff and volunteers will be associated with greater confidence in elections administration compared to announcements that do not mention military veterans.

I considered that veterans would elicit high degrees of approval and support for whatever cause or issue presented. For example, I'd expect high approval and support for recruiting veterans to baseball teams just as much as for recruiting them to work as election staff. Likewise, asking participants whether they supported programs intent on recruiting anyone to work as election staff would garner support regardless of the target group. Considering that support and admiration for veterans is generally high among the population [@kleykamp2023], then discerning the impact on one's confidence would require a survey experiment designed to determine whether publicized recruitment efforts targeting veterans as a group would have any special effect on confidence in elections administration.

Since the results and events following the 2020 election loomed large in anticipation of the U.S. 2024 general elections, it is reasonable to take into account beliefs about legitimacy of the 2020 election. This group who denies the legitimacy of the 2020 election results were possibly the most distrusting of the 2024 election. However, given that veteran service members are held in high regard, generally speaking, announcing that veterans are being actively recruited to work as election staff and volunteers is expected to positively influence confidence in elections among those who remained firm in the belief that the 2020 election results were illegitimate. I take into account that legitimacy beliefs about the 2020 election and expectations of electoral fraud are related to partisanship and include partisanship as a control in the analysis.

In addition, I expect that both expectations of electoral fraud and concerns for voter safety would lessen (shrink) when presented with information that veterans are being actively recruited to work and volunteer in election offices.

Finally, prior research has shown that confidence that one's own vote would be counted as intended has been much stronger than confidence that votes nationwide would be treated likewise [@sances2015; @stewart2022]. In other words, the public is more trusting of elections closer to home. Taking into account that confidence is lower for elections conducted outside of one's local area, then information that veterans are being recruited to help conduct elections in locations beyond one's locale may close this gap between confidence in elections within one's local area and confidence in elections elsewhere. Although it may be difficult to discern whether recruiting veteran service members would improve confidence in elections near one's local area in light of such hometown favoritism, a positive effect on confidence in elections held beyond one's local area---where confidence is already expectedly lower---can help discern the impact of the group alone. Therefore, any observed disparity of confidence in elections (i.e., the "confidence gap") between elections within one's local area and elections beyond will be smaller when people are presented with information about election official recruitment efforts of veteran service members.

## Experiment Design and Survey Measures

To test the theory that publicized efforts to recruit veterans to work as election staff and volunteers would improve confidence in elections and ease insecurity, a recent experiment was embedded in a survey developed and conducted by the Center for Democracy and Civic Engagement (CDCE) at the University of Maryland. The survey was fielded from August 29th, 2024 to September 18th, 2024 on a non-probability sample of 1,287 U.S. citizens 18 years of age or older. Respondents were randomly split into either the treatment ($n =$ 650) control ($n =$ 637) conditions[^4].

Survey participants read either a treatment or control vignette, which was a fabricated news article about efforts in Maricopa County, AZ to recruit election staff and volunteers for the 2024 general election. The treatment vignette referred to a program designed to recruit veterans and their family members and describes an interviewee "Jordan Braxton" as an Army veteran. The control vignette simply omitted any mention of veterans and their family members, and didn't describe "Jordan Braxton" as an Army veteran. Beyond those small differences and the headlines, the article vignettes are identical. Therefore, effects can be attributed to the information about veterans in the treatment vignette[^5].

It should be noted that Maricopa County, AZ was chosen as the setting of the story in the vignette due to the increased scrutiny levied toward election administration there after the 2020 election [@giles2021; @maricopacountyelectionsdepartment2022]. Because of this, treatment effects are potentially limited or constrained to attitudes concerning the location specified in the vignette. A block of survey items asked specifically about Maricopa County, AZ, followed by an identical block of items that asked the same questions about one's local area. Later, I compare and discuss results between the items pertaining to different settings.

@@

Note that, except for the item assessing voter confidence (i.e., votes counted as voters intend), most other question items from the 2022 SPAE inquire into the voter experience after the election has occurred. The items in this study, however, all inquire into confidence over one's expectations in anticipation of upcoming elections. In addition, other survey questionnaires ask similar questions that specifically contrast between a respondent's own local community, their state, and nationwide; this is in addition, or complementary to, items that distinguish between local area officials, state election officials, state government, and the U.S. federal government. The survey of this study, instead, focused on attitudes regarding election workers such as election officials, staff, and volunteers specific to the location mentioned in the vignette---Maricopa County, AZ---as well as specific to one's local area[^6].

The survey included items that assessed both trust in elections administration and expectations of electoral fraud, referred to simply as *trust* and *distrust*, respectively. Responses from these two sets of items were utilized to construct the primary dependent variable, *confidence in elections*.

Prior to the reading the treatment or control vignette, participants responded to distinct survey items assessing one's favorability to local election officials ("In general, how favorable or unfavorable is your impression of local election officials?") and legitimacy of the 2020 election results ("Regardless of whom you supported in the 2020 election, do you think Joe Biden's election as president was legitimate, or was he not legitimately elected?"). The former presented five response options from "Strongly favorable" to "Strongly unfavorable"; the latter item presented a binary of either "Legitimate" or "Not Legitimate".

The conventional multiple item assessment of partisanship was included and used to construct a variable of partisanship consisting of three categories: Democrat, Republican, and Independent. In this case, Independents are "true" Independents in that they do not identify with either political party and do not "lean" toward either Republican or Democratic parties. Self-described "Independent leaners" are re-coded as falling into the partisan category in which they lean e.g., Independents who expressly report that they lean Republican are denoted as Republican.

Finally, a pair of survey items assessed concerns for potential violence and confidence in voter safety. The pair was asked once in regard to Maricopa County, AZ and again with respect to one's local area[^7].

### Trust and Distrust in Elections

A set of five survey items measured public trust in elections by inquiring into the degree of confidence that votes will be counted as voters intend, confidence that electoral systems are secure from technological threats, perceived commitment of election staff, confidence that outcomes will be fair, and confidence that the voting process will be fair. Each item in this series presented four response options from "Not at all \[confident/committed\]", "Not too \[confident/committed\]", "Somewhat \[confident/committed\]", and "Very \[confident/committed\]". In total, five survey items measure different aspects of trust in elections administration. The items measuring confidence in elections include:

Another series of five survey items captured an individual's level of *distrust* in elections administration based on the extent to which they *expect electoral fraud* to occur. These items were prefaced with the question, "How likely do you think any or all of the following will happen during this year´s elections in \[Maricopa County, AZ/ your local area\]?" Each item presented four response options: "Not likely at all", "Not too likely", "Somewhat likely", and "Very likely".

```{r}
#| label: tbl-items
#| tbl-cap: "Survey Items for Trust and Distrust in Elections"

trust_txt <- df |> select(q19, q20, q21, q22, q24) |>
  labelled::set_variable_labels(
    q21 = "How commited do you think election staff and volunteers in Maricopa County, AZ will be to making sure the elections are fair and accurate?"
  ) |> 
  labelled::generate_dictionary() |> 
  dplyr::as_tibble() |> 
  select(Trust=label)

distrust_txt <- df |> select(q28_1:q28_5) |>
  labelled::generate_dictionary() |> 
  dplyr::as_tibble() |> 
  select(Distrust=label)

conf_items_text <- trust_txt |> 
  mutate(Distrust = pull(distrust_txt))

kableExtra::kbl(conf_items_text, booktabs = T,
                col.names = c("Trust", "Distrust"),
                linesep = "\\addlinespace") |> 
  kableExtra::kable_styling(full_width = T,
                            font_size = 8,
                            latex_options = c("HOLD_position")) |> 
  kableExtra::column_spec(column = 1:2, width = "8cm")

Similar to the items capturing trust in elections (@tbl-items), I computed a summated score for each respondent reflecting the extent of their distrust in elections.

Upon construction of the two summated scales of *trust* and *distrust*, internal consistency reliability coefficient $\alpha$ were estimated for both and found to be adequate ($\alpha$ \> 0.8); the trust in elections scale $\alpha$ = 0.93 and the distrust scale $\alpha$ = 0.86. In other words, the items in each of the scales were intercorrelated well enough for each to compose a single scale presuming a unidimensional construct for each.

I expected item responses from these two scales to be inversely correlated among the sample, but not mutually exclusive. Indeed, this is what I find. Polychoric correlations between the trust and distrust items negatively correlate as expected\[\^See Appendix C\], as does the correlation between the two sum score scales. Given the ordinal nature of the variable items, I conducted a Spearman's rank correlation test [@spearman1907] and found a negative correlation of $\rho = -0.49$ ($95\%$ CI \[-0.53, -0.44\]; Kendall's $\tau = -0.39$) between scores on the two scales for the AZ items (For the local item scale score correlation: $\rho = -0.52$, $95\%$ CI \[-0.56, -0.48\]; Kendall's $\tau = -0.39$). The negative correlation between the trust and distrust items and scores makes intuitive sense.

## Data Specification

The primary outcome of interest concerns public confidence in elections administration compensated by one's distrust in elections, i.e., their expectations of electoral fraud. Because trust and distrust are assessed by multiple Likert items, conventional methods call for combining the items of each set into separate Likert scales (also known as summated rating scales). Since the item response options are not symmetric---i.e., not bipolar with contrasting positive and negative response options to include a neutral middle option---then the *trust in elections* scale could only be understood as a measure of trust in the positive use of the term (i.e., unipolar). Likewise, the expectation of fraud items fit to a scale that corresponds to a unipolar measure of confident *distrust*, so to speak. However, many of the items comprising the trust scale directly oppose items capturing distrust. For instance, one item from the trust scale has the respondent report their confidence that votes will be counted as voter's intend, whereas a separate item from the distrust scale has the the respondent report their expectation that many votes will not be counted. Positive responses to these two items directly contradict each other, and would thus cancel out.

The relationship is that trust and distrust represent opposing degrees of confidence along the same spectrum. However, despite the inherent contrast, one can hold both positive degrees of trust and distrust, feeling similarly confident in both respects. Accordingly, equivalent degrees of trust and distrust cancel out, rendering one relatively more *insecure* about their expectations of the future despite the fact that they may feel confident in either direction. When trust and distrust are considered as reflective of one's confidence about future expectations, then both can be placed along the same spectrum. Therefore, a lack of confidence denotes insecurity.

Something may cause a person to have greater expectations that election fraud will occur, which in turn will always lower their trust in elections administration. In contrast, a person may feel a bit more confident in the integrity of elections, but this confidence is constrained by the extent to which they expect election fraud to occur. This means that a person's baseline level of confidence in elections admin must take into account their present degree of distrust---i.e., their expectations of electoral fraud.

Those who are, or are close to being, equally confident in both directions would be the most *insecure* about their expectations. Having no trust in elections and no expectation that electoral fraud will occur is functionally equivalent to having equal degree of both. Quite literally, this reflects an uncertain feeling of "it could go one way or the other" with respect to the integrity of elections administration. Operationally, this would reflect a middling point on a scale of *confidence*.

In light of this, I reverse coded positive expectations of electoral fraud and added responses from the distrust scale to the trust in elections scale. That is, item responses on the trust scale ranged from 0 ("Not confident at all") to 3 ("Very confident), whereas responses on the distrust scale ranged from 0 ("Not likely at all") to -3 ("Very likely"). Combining responses in this way resulted in an overall measure of *confidence in elections* to serve as the dependent variable. On this scale, positive values reflect higher confidence in elections, negative values reflect greater distrust, and values closer to zero correspond to an individual's degree of insecurity[^8].

The advantage is that this permits a better test of the treatment effect as theorized. Of course, it is easy enough to examine the effect of the treatment on both trust and distrust separately. However, the theory presents a hypothesis that the treatment will impose influence upon both variables *concurrently and in a positive direction* in light of the relationship between trust and distrust. Confidence increases insofar as trust in elections increases and distrust decreases; since distrust detracts from trust, then an accurate measure of confidence must take that constraint into account.

@@

I first review responses to the survey items capturing trust and distrust, respectively. When distinguished by experiment condition, trust in elections admin is noticeably higher for those who read the treatment vignette. The difference is more pronounced for some items over others. That being said, there is a definite distinction between items pertaining to Maricopa County, AZ compared to items that pertain to a respondent's local area. In fact, the treatment appears to have hardly made a difference when the items concerned one's local area.

```{r}
#| label: fig-trustlikert
#| fig-cap: "Trust in Election Administration by Experiment Condition"

# Likert plot of trust and confidence in election admin comparing treatment and
# control groups
trust.likert.treat.az <- df |>
  # relabel factor labels for plotting
  mutate(across(
    c(q19, q20, q21, q22, q23, q24, q30, q31, q32, q33, q34, q35),
    ~ pewmethods::fct_case_when(
      .x %in% c("Not at all confident", "Not at all committed") ~ "Not at all confident",
      .x %in% c("Not too confident", "Not too committed") ~ "Not too confident",
      .x %in% c("Somewhat confident", "Somewhat committed") ~ "Somewhat confident",
      .x %in% c("Very confident", "Very committed") ~ "Very confident",
      TRUE ~ NA
    )
  )) |>
  ggstats::gglikert(
    include = c(q19, q20, q21, q22, q24),
    y = 'group',
    facet_rows = vars(.question),
    symmetric = F,
    labels_size = 2.5,
    add_totals = T
  ) +
  ggplot2::scale_fill_grey(start = 0.5, end = 0.1)+
  # ggplot2::scale_fill_brewer(palette = "Greys")+
  labs(subtitle = str_wrap("Maricopa County, Arizona", width = 95)) +
  theme_bw() +
  theme(
    text = element_text(family = "Palatino Linotype"),
    legend.position = 'bottom',
    axis.text.x = element_blank(), # remove percentage text along x-axis
    strip.text.y.right = element_blank()
  )
